---
title: "BIOMI668 Final Project"
output: html_document
date: "2024-05-01"
---
```{r}
library(ggplot2)
```

## Uploading and trimming the data

```{r}

#Upload data
#Data set contains samples as the columns and genes identified as the rows
count_data = read.csv("gene_counts_data.csv")

#Mapping file
map = read.csv("sohl_map.csv")


#Removing rows that only contain zeros
trimmed_df <- count_data[rowSums(count_data[,-c(1, 2)]) != 0, ]

#Write it to a csv
write.csv(trimmed_df, file = "trimmed_count_data.csv")

#Data set without the gene length values, so these values don't affect the CurvCut code
dataset = trimmed_df[,-c(2)]
write.csv(dataset, file = "ultratrim_count_data.csv")

```

### The dataset, "ultratrim_count_data.csv", is ran through CurvCut. 

## Uploading the data that CurvCut created

CurvCut was created to proved a mathematical backing to eliminating sparse features in a count tables.

```{r}
#Upload the new data set from CurvCut

ultratrim_data_zf = read.csv("ultratrim_count_datatable.zerofiltered.csv", row.names = 1)
ultratrim_data_zf = ultratrim_data_zf[,-c(1)]

```

## Need to transpose the data set for plotting, so the samples are on the rows and the gene ids are the columns


```{r}
#Need to transpose the data set for plotting, so the samples are on the rows and the gene ids are the columns
rownames(ultratrim_data_zf)<-ultratrim_data_zf$X
ultratrim_data_zf$X<-NULL

dt_transpose = t(ultratrim_data_zf)
dt_transpose = as.data.frame(dt_transpose)
#head(dt_transpose)

#Assign the first row as column names
colnames(dt_transpose) <- dt_transpose[1, ]

#Remove the first row
dt_transpose <- dt_transpose[-1, ]

```

## Perform CLR Transformation

```{r}

dt_zerorep <- dt_transpose

#Convert the data frame to numeric, otherwise the function won't work on characters
#Set the sample names to row names because the conversion of characters to numeric changes the samples names to numbers
sample_names <- rownames(dt_zerorep)
dt_zerorep <- as.data.frame(sapply(dt_zerorep, as.numeric))

#Replace the zeros with a small number
dt_zerorep[dt_zerorep == 0] <- 0.001

#Re-add the sample names back into the data
rownames(dt_zerorep) <- sample_names

#Set the clr function
clr <- function(x) sweep(log(x), 1, rowMeans(log(x)), "-")

#Now run the transformation and assign to a new data frame
transformed_data <- data.frame(clr(dt_zerorep))

```

## Run a PCA on the clr-transformed feature table

```{r}
#PCA with CLR transformed data
pca <- prcomp(transformed_data)

#Summary
summary_pca <- summary(pca)
summary_pca

summary_pca$importance

```
## Finding the most important geneid for the first component

```{r}

pca_loadings <- data.frame(pca$rotation)

pca_loadings_PC1 <- pca_loadings[order(-pca_loadings$PC1), ]

#Show the top 10 values for PC1
#I just chose the 10 top values, so all 27411 variables don't knit
head(pca_loadings_PC1, 10)


``` 
TFPI2 gene is the most important for PC1.


## Finding the most important geneid for the second component

```{r}
pca_loadings_PC2 <- pca_loadings[order(-pca_loadings$PC2), ]

#Show the top 10 values for PC2
#I just chose the 10 top values, so all 27411 variables don't knit
head(pca_loadings_PC2, 10)


```
ADHIC gene is the most important for PC2. 

## Add the variables for SampleGroup, TumorType, and SampleNumber to the PCA data frame


```{r}

pca_df = as.data.frame(pca$x)
pca_df$Sample.Number <- map$Sample.Number
pca_df$Sample.Group <- map$Sample.Group
pca_df$TumorType <- map$Tumor.Type


```

## Plot PC1 against PC2 based on the sample groups of mutations: R132H, R132Q, and WT

```{r}
p_PCA_group <- ggplot(data = pca_df) + 
  geom_point(aes(x = PC1, y = PC2, color = Sample.Group)) + 
    stat_ellipse(geom = "polygon",aes(x = PC1, y = PC2, fill = Sample.Group, color = Sample.Group), alpha = 0.18) +
  theme_classic() +
  labs(title= "PCA") +
  theme(text = element_text(size = 15)) 
p_PCA_group


```


## Plot PC1 against PC2 based on the sample type of tumors: HT1080 and U87MG

```{r}
p_PCA_tumor_type <- ggplot(data = pca_df) + 
  geom_point(aes(x = PC1, y = PC2, color = TumorType)) + 
    stat_ellipse(geom = "polygon",aes(x = PC1, y = PC2, fill = TumorType, color = TumorType), alpha = 0.18) +
  theme_classic() +
  labs(title= "PCA") +
  theme(text = element_text(size = 15)) 
p_PCA_tumor_type

```



